---
title: COVID-19 Transmission
description: Virus outbreak transmission analysis based on graph visualization, using visNetwork package.
author: ''
date: '2020-03-24'
slug: covid-19-transmission
categories: []
tags:
  - portofolio
  - dataplay
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
Sys.setlocale("LC_TIME", "Ind")
library(RJSONIO)
library(tibble)
library(dplyr)
library(stringr)
library(tidyr)
library(lubridate)
library(visNetwork)
```

# Preface

This content is created for educational purposes, shout out to [louislugas](https://github.com/louislugas/) from GitHub for the content inspiration. Check out [his awesome web application](https://louislugas.github.io/covid_19_cluster/index.html) for COVID-19 transmission network analysis. This article will focus on breaking down building blocks in creating network analysis taking on the most recent COVID-19 pandemic outbreak in Indonesia. There are exponentially rising amount of suspects and patients on this matter and a strategic approach in containing the spread will be required in decision makings in this country. This article, will not tackle the policy-maker perspective, instead introducing the potentially useful tools for some parties.

# Graph Theory

The study of graph are introduced as mathematical structure used to model a certain relationships between entities. Graph is applicable in many areas and expertises, but a simple graph contain 2 main components:

* Nodes (or Vertices): Is the entity represented in a graph, in our case, the node will be COVID-19 patients recorded so far
* Edges: The edges, describe the relation between nodes

To acquire the data, I will be using JSON provided by [louislugas](https://github.com/louislugas/), used also in his application:
 
```{r}
json_cases <- fromJSON("https://raw.githubusercontent.com/louislugas/louislugas.github.io/master/covid_19_cluster/json/kasus-corona-indonesia.json")
cases_df <- as.data.frame(json_cases$nodes) %>% 
  t() %>% 
  as.data.frame() %>% 
  glimpse()
```

> Note: JSON is an unstructure data types, while data frame is a set of variables (columns) with a given amount of observations (rows). Sometimes, JSON is not easily convertable to a tabular data type like data frame. Make sure you understand JSON's data structure before making a table out of it.

## The Nodes

Well, since most of them has an intuitive column name, we wouldn't need a lot of exploration in understanding the data. The `penularan` column, however, seems to be a free-writing column that should be further inspected:

```{r}
set.seed(10)
cases_df %>% 
  select(penularan) %>% 
  sample_n(10)
```

The `Tidak Diketahui` value seems to be really common, and might not be too useful to analyze if it holds no connection. So we'll filter this out:

```{r}
known_cases <- cases_df %>% 
  filter(penularan != "Tidak diketahui") %>% 
  glimpse()
```

The cases is now down to 36 cases that might (or might not be) connected to each other. Let's take a look at the `penularan` column once again:

```{r}
set.seed(10)
known_cases %>% 
  select(penularan) %>% 
  sample_n(10)
```

A bit of common pattern we're seeing here:

* Relation with other case seems to be indetified using the `#<a number>` value
* The word `Sehabis` is a strong indicator of an imported case

Now to create a case - when logic, we'll be using `dplyr`'s `case_when()` to map each of the string pattern using regex:

```{r}
relation_df <- known_cases %>% 
  mutate(
    relation = case_when(
      str_detect(penularan, "[0-9]+") ~ str_extract(penularan, "#[0-9]+"),
      str_detect(penularan, "Sehabis") ~ "Imported case",
    )
  ) %>% 
  select(id, provinsi, umur, gender, pengumuman, penularan, status, wn, rs, relation)
```

Now glimpsing if there's anything missed out from the given patterns:

```{r}
relation_df %>% 
  filter(is.na(relation)) %>% 
  select(id, penularan)
```

So 3 more uncommon pattern:

* The Diamond Princess incident seems to be used as an identifier as well
* `WNA` indicates a foreigner which will be assumed to be the same as `Imported Case`
* A very long description that might (or might not have) explains the transmission, we'll ignore that for now

So rewriting our `relation_df` codes and adding a few informative information extraction from other columns:

```{r}
nodes <- known_cases %>% 
  mutate(
    relation = case_when(
      str_detect(penularan, "[0-9]+") ~ str_extract(penularan, "#[0-9]+"),
      str_detect(penularan, "Sehabis") ~ "Imported case",
      str_detect(penularan, "Diamond Princess") ~ "Diamond Princess",
      str_detect(penularan, "WNA") ~ "Imported case"
    ),
    # coercing to date type for date calculation
    pengumuman = dmy(pengumuman),
    time_since = difftime(Sys.Date(), pengumuman, units = "d")
  ) %>% 
  select(id, pengumuman, provinsi, umur, status, wn, rs, relation, time_since, penularan)
```

Now there here goes our nodes data frame:

```{r echo=FALSE}
knitr::kable(nodes %>% head())
```

## The Edges

To acquire each case's connection, we'll be simply extracting the information from our node since it has come with a handy identifier using `relation` and `id` columns. Starting from our `relation` column, we'll be extracting only with those that have relation between cases, means those that has the `#<a number>` value in it:

```{r}
edges <- relation_df %>% 
  select(relation, id) %>% 
  mutate(
    relation = str_extract(relation, "[0-9]+"),
    relation = as.numeric(relation)
  ) %>% 
  drop_na() %>% 
  rename(
    from = relation,
    to = id
  )
```

And here goes our edges data frame:

```{r echo=FALSE}
knitr::kable(edges %>% head())
```

Now before going any further, the first case seems to be showing a weird connection to the 24th case, if the timeline is correct, the first case shouldn't be infected from the latter cases. Now let's inspect the original dataset:

```{r}
known_cases %>% 
  filter(id == 1) %>% 
  select(penularan)
```

The `penularan` column describes how it is related to `kasus #24`, things to note is, it took me some time to understand this anomaly, which I found to be really useful to just Google the things I find off. Then I came accross the [following article](https://www.cnnindonesia.com/internasional/20200229110333-106-479291/tiga-kasus-baru-corona-di-malaysia-satu-pernah-ke-indonesia) that explains that the `kasus #24` is indicating to Malayasia's case instead of Indonesia's. Means, we can annull the first edges before going to visualization:

```{r}
edges <- edges[-1,]
```

# Using `visNetwork`

There are indeed load of packages for graph visualization. For this exercise we'll be using `visNetwork` and please note that I always encourage people to aim for understanding the working principles and ideas - not the idiosyncrasies. For the `visNetwork` to work, we'll be needing some components:

* `nodes` data frame that consist of:
  * Id
  * The color
  * Size
  * Shape
  * Label
  * Hover text
* `edges` data frame consisting of `to` and `from` that correspond to `id` column in `nodes`

Some things I want to configure based on the given set up is:

* The status of the patient will be color-coded

```{r}
levels(nodes$status)
```

* The shape will be diamond if it is an imported case and square if not
* The size will be based on days passed since announcement
* The hover text will show: age, hospital location, and details

Since we'll be needing some value mapping, we'll split it into 2 parts: data preparation and visualization

## Data Mapping

`visNetwork` uses column names to map each value to its designated component. Simply put:

* Use `id` for node's ids
* Use `label` for node's names
* Use `shape` for node's shapes
* Use `color` for node's colors
* Use `title` for node's hover text
* Use `value` for node's size


```{r}
cleaned_nodes <- nodes %>% 
  mutate(
    label = paste0("Case #", id),
    color = case_when(
      status == "Sembuh" ~ "green",
      status == "Dalam Perawatan" ~ "orange",
      status == "Meninggal" ~ "red",
      TRUE ~ "grey"
    ),
    shape = case_when(
      str_detect(relation, "Imported case") ~ "diamond",
      TRUE ~ "square"
    ),
    # The hover text uses HTML formatting so <br> means break or new line
    title = paste("Umur:", umur, "<br>", 
                  "Lokasi:", rs, "<br>", 
                  "Keterangan:", penularan),
    value = time_since
  ) %>% 
  select(id, label, shape, color, title, value)
```

Now to take a glimpse on the cleaned data frame:

```{r echo=FALSE}
knitr::kable(cleaned_nodes %>% head())
```

Now all seems good, and the edges data frame also does not seems like in need of any wrangling and cleansing:

```{r echo=FALSE}
knitr::kable(edges %>% head())
```

## Graph Visualization

Last is to just pass in our `nodes` and `edges` along with some configuration. For a complete documation for the API, please visit the [official documentation](https://datastorm-open.github.io/visNetwork/). The configuration I am using a simple tweak for aesthetic purposes and layouting for better view:

```{r}
visNetwork(cleaned_nodes, edges) %>% 
   visNodes(shadow = list(enabled = TRUE, size = 10), 
            font = list(size = 24)) %>% 
  visHierarchicalLayout(direction = "LR", levelSeparation = 300)
```

Now try to play around with the visualization above. Note that you can scroll to zoom and h See if you can replicate the same (or different!) visualization by understanding how graph works!